---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 20 @ 11:59PM
author: "Your Name and UID"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

```{r}
library(tidyverse)
library(tidymodels)
library(gtsummary)
library(naniar)  # Visualize missing data
library(ggplot2)
library(recipes)
```

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 


1. Data preprocessing and feature engineering.

```{r}
# Load Data
cohort_data <- readRDS("mimic_icu_cohort.rds")

# Select Features & Target
mimiciv_icu_cohort <- cohort_data |>
  select(
    subject_id, hadm_id, stay_id,
    
    los_long, # Target variable (ICU stay > 2 days)
    
    # Demographics
    gender, age_intime, marital_status, race,
    
    # ICU admission
    first_careunit,
    
    # Lab measurements (Last before ICU)
    bicarbonate, chloride, creatinine, glucose, hematocrit, 
    potassium, sodium, wbc,
    
    # Vital measurements (First in ICU)
    heart_rate, non_invasive_blood_pressure_diastolic, 
    non_invasive_blood_pressure_systolic, respiratory_rate, 
    temperature_fahrenheit
  ) |>
  # Convert categorical variables to factors
  mutate(
    gender = as.factor(gender),
    marital_status = as.factor(marital_status),
    race = as.factor(tolower(race))
  ) |>
  drop_na(los_long) |>
  print(width = Inf)
```

```{r}
# Check for missing values
mimiciv_icu_cohort |> tbl_summary(by = los_long)
```

There are missing values in `marital_status`, lab measurements, and vital measurements.
Missing values are visualized before deciding how to handle them.

```{r}
vis_miss(mimiciv_icu_cohort |>
           select(-subject_id, -hadm_id, -stay_id, -los_long, -gender, 
                  -age_intime, -race, -first_careunit), 
         warn_large_data = FALSE)
```

```{r}
numeric_cols <- select(mimiciv_icu_cohort |>
                         select(-subject_id, -hadm_id, -stay_id, -los_long, 
                                -gender, -age_intime, -race, -first_careunit), 
                where(is.numeric))  # Select numeric columns

# Loop through each numeric column and create a histogram
for (col in names(numeric_cols)) {
  p <- ggplot(mimiciv_icu_cohort, 
              aes_string(x = col)) +
    geom_histogram(binwidth = 10, fill = "skyblue", 
                   color = "black", alpha = 0.7) +
    labs(title = paste("Histogram of", col), x = col, y = "Frequency") +
    theme_minimal()
  print(p)
}
```

Mean imputation should be used for normally distributed variables, `bicarbonate`, `chloride`, `hematocrit`, `sodium`, and `heart_rate`.

Median imputation should be used for skewed variables, `creatinine`, `glucose`, `potassium`, `wbc`, `non_invasive_blood_pressure_diastolic`, `non_invasive_blood_pressure_systolic`, `respiratory_rate`, and `temperature_fahrenheit`.

```{r}
ggplot(mimiciv_icu_cohort, aes(x = marital_status)) +
  geom_bar() +
  labs(title = "Bar plot of Age at ICU Intime by LOS Long",
       x = "Marital Status") +
  theme_minimal()
```

KNN imputation should be used for categorical variables, `marital_status`.

2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r}
#| eval: false
set.seed(203)

# sort
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )
data_split
```

```{r}
mimiciv_icu_cohort_train <- training(data_split)
dim(mimiciv_icu_cohort_train)
mimiciv_icu_cohort_test <- testing(data_split)
dim(mimiciv_icu_cohort_test)
```

3. Train and tune the models using the training set.

**Logistic regression with enet regularization:**


```{r}
rec <- recipe(los_long ~ ., data = mimiciv_icu_cohort) |>
  # Mean inputation for normal variables
  step_impute_mean(bicarbonate, chloride, hematocrit, sodium, heart_rate) |>
  
  # Median imputation for skewed variables
  step_impute_median(creatinine, glucose, potassium, wbc, 
                     non_invasive_blood_pressure_diastolic, 
                     non_invasive_blood_pressure_systolic, 
                     respiratory_rate, temperature_fahrenheit) |>
  
  # KNN imputation for categorical variables
  step_impute_knn(marital_status)

# Apply recipe to the data
prepped_data <- prep(rec, training = mimiciv_icu_cohort)
mimiciv_icu_cohort <- bake(prepped_data, new_data = mimiciv_icu_cohort)

head(mimiciv_icu_cohort)
```

```{r}

# Check missing values after imputation
colSums(is.na(mimiciv_icu_cohort))
```

4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?
